---
output:
  knitrBootstrap::bootstrap_document:
    theme: readable
    highlight: zenburn
    theme.chooser: TRUE
    highlight.chooser: TRUE
  html_document:
    toc: true
    highlight: zenburn
---

```{r setup}
library(knitr)

library(ggplot2)
library(reshape)
library(DESeq2)
library(genefilter)
library(CHBUtils)
library(gtools)
library(gridExtra)
library(devtools)
library(dplyr)
library(isomiRs)

knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png", fig.width=6,fig.heigh=6,
               cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='', bootstrap.show.code=FALSE)

root_path = "$path_abs"
root_file = paste0(root_path,"/report/")

condition = "condition"
```

```{r render,eval=FALSE, echo=FALSE}
library(rmarkdown)
library(knitrBootstrap)
# render("report.rmd")
```

```{r load}
setwd(root_path)
files = read.table("$path_summary", sep=",",header=T,colClasses = "character")

samples = files[,"sample_id"]

names_stats = files[,"size_stats"]
names(names_stats) = samples

groups = files[,"group"]
names(groups) = samples

summarydata = data.frame(row.names=samples,samples=samples,group=groups)
design <- data.frame(row.names=files$sample_id, condition=files$group)

```

# Exploratory analysis

In this section we will see descriptive figures about quality of the data, 
reads with adapter, reads mapped to miRNAs, reads mapped to other small RNAs. 

## size distribution

After adapter removal, we can plot the size distribution of the small RNAs.

```{r adapter,fig.width=10}
tab = data.frame()
for (sample in samples){
    d = read.table(names_stats[sample], sep=" ")
    tab = rbind(tab, d %>% mutate(sample=sample, group=groups[sample]))
}


reads_adapter = tab %>% group_by(sample, group) %>% summarise(total=sum(V2))
ggplot(reads_adapter, aes(x=sample,y=total,fill=group)) +
    geom_bar(stat="identity", position = "dodge") +
    ggtitle("total number of reads with adapter") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


ggplot(tab, aes(x=V1,y=V2,fill=sample)) +
    geom_bar(stat="identity", position = "dodge") +
    facet_wrap(~group, nrow=2)+
    ggtitle("size distribution") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## miRNA

### total miRNA expression annotated with mirbase

```{r mirna,results='hide'}
mi_files = files[,"miraligner"]
row.names(design) = samples

obj <- loadIso(files = mi_files, design = design ,header = T, cov = 1)
```


```{r mirna-mirbase}
ggplot( data.frame(sample=colnames(obj@counts), total=colSums(obj@counts))) +
    geom_bar(aes(x=sample,y=total), stat='identity')+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
mirna_step <- as.data.frame(colSums(obj@counts))
```

### Distribution of mirna expression

```{r depth}
ggplot(melt(obj@counts)) +
    geom_boxplot(aes(x=X2,y=value))+
    scale_y_log10()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

### cumulative distribution of miRNAs

We can see this time we detect much more miRNAs. The saturation is near to 200.

```{r cum}

cs <- as.data.frame(apply(obj@counts,2,cumsum))
cs$pos <- 1:nrow(cs)

ggplot((melt(cs,id.vars = "pos")))+
    geom_point(aes(x=pos,y=value,color=variable))+
    scale_y_log10()
```

## Others

### classification

```{r cluster}
clus <- read.table(files[1,"clusters"],header=T,sep="\t",row.names=1)
ann <- clus[,1]
clus_ma <- clus[,2:ncol(clus)]
clus_ma = clus_ma[,row.names(design)]
```

The major contribution are repeats and rRNA, with variation in the percentage. 

```{r cluster_type,eval=FALSE}
rRNA <- colSums(clus_ma[grepl("rRNA",ann) & grepl("miRNA",ann)==F,])
miRNA <- colSums(clus_ma[grepl("miRNA",ann),])
tRNA <- colSums(clus_ma[grepl("tRNA",ann) & grepl("rRNA",ann)==F & grepl("repeat",ann)==F & grepl("miRNA",ann)==F,])
rmsk <- colSums(clus_ma[grepl("repeat",ann) & grepl("rRNA",ann)==F & grepl("miRNA",ann)==F,])
total <- colSums(clus_ma)

dd <- data.frame(samples=names(rRNA),
                 rRNA=rRNA,
                 miRNA=miRNA,
                 tRNA=tRNA,
                 rmsk=rmsk,
                total=total)
ggplot(melt(dd)) +
    geom_bar(aes(x=samples,y=value,fill=variable),
             stat='identity',
             position="dodge")+
    scale_fill_brewer(palette = "Set1")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

dd_norm = dd
dd_norm[,2:5] = sweep(dd[,2:5],1,dd[,6],"/")
ggplot(melt(dd_norm[,1:5])) +
    geom_bar(aes(x=samples,y=value,fill=variable),
             stat='identity',
             position="dodge")+
    scale_fill_brewer(palette = "Set1")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
    labs(list(title="relative proportion of small RNAs",y="% reads"))
```


# DE

I did differential expression of all possible comparison using the previous normalization.
For each DE I tried, I plot Dispersion plot that shows the relationship between
standard deviation expression and average expression for each miRNA.


```{r de-setup}
library(DESeq2)
library(DEGreport)
library(vsn)
```

```{r deseq2-handler}

filter_handle <- function(res){
    res_nona <- res[!is.na(res$padj),]
    keep <- res_nona$padj < 0.1 
    res_nona[keep,]
}

handle_deseq2 = function(dds, summarydata, column) {
  all_combs = combn(levels(summarydata[,column]), 2, simplify=FALSE)
  all_results = list()
  contrast_strings = list()
  for(comb in all_combs) {
    contrast_string = paste(comb, collapse="_vs_")
    contrast = c(column, comb)
    res = results(dds, contrast=contrast)
    res = res[order(res$padj),]
    all_results = c(all_results, res)
    contrast_strings = c(contrast_strings, contrast_string)
  }
  names(all_results) = contrast_strings
  return(all_results)
}

plot_MA = function(res){
    for(i in seq(length(res))) {
        plotMA(res[[i]])
        title(paste("MA plot for contrast", names(res)[i]))
    }
}

plot_volcano = function(res){
    for(i in seq(length(res))) {
        stats = as.data.frame(res[[i]][,c(2,6)])
        p = volcano_density_plot(stats, title=names(res)[i], lfc.cutoff=1)
        print(p)
    }
}

do_de = function(raw, summarydata, condition){
    dss = DESeqDataSetFromMatrix(countData = raw[rowMeans(raw)>3,],
                       colData = summarydata,
                       design = ~ condition)
    dss = DESeq(dss)
    plotDispEsts(dss)
    dss
}

do_norm = function(dss, root_path, prefix){
    rlog_ma = assay(rlog(dss))
    count_ma = counts(dss, normalized=TRUE)
    raw = counts(dss, normalized=FALSE)
    fn_log = paste0(root_file, prefix, "log_matrix.txt")
    write.table(rlog_ma,fn_log,sep="\t")
    fn_count = paste0(root_file, prefix, "count_matrix.txt")
    write.table(count_ma,fn_count,sep="\t")
    fn_raw = paste0(root_file, prefix, "raw_matrix.txt")
    write.table(count_ma,fn_raw,sep="\t")
}

```

## mirna

```{r de}
dds = do_de(obj@counts, design, condition)
```

```{r tables}
do_norm(dds, root_path, "mirna_")
```

### MA-plots
  
```{r DESeq-output, results='asis'}
all_results = handle_deseq2(dds, design, condition)
plot_MA(all_results)
```

### Volcano-plots

```{r DESeq-volcano}
plot_volcano(all_results)
```

## isomir

```{r de-iso}
dds = do_de(countsIso(obj, ref=T, iso5=T, iso3=T, add=T, mism =T)@counts, design, condition)
```

```{r tables-iso}
do_norm(dds, root_path, "isomirs_")
```

### MA-plots
  
```{r DESeq-output-iso, results='asis'}
all_results = handle_deseq2(dds, design, condition)
plot_MA(all_results)
```

### Volcano-plots

```{r DESeq-volcano-iso}
plot_volcano(all_results)
```

## clusters

```{r de-c}
dds = do_de(clus_ma, design, condition)
```

```{r tables-c}
do_norm(dds, root_path, "clusters_")
```

### MA-plots
  
```{r DESeq-output-c, results='asis'}
all_results = handle_deseq2(dds, design, condition)
plot_MA(all_results)
```

### Volcano-plots

```{r DESeq-volcano-c}
plot_volcano(all_results)
```



